---
title: "Chapter 2 The Very Basics"
author: "Hendra Bunyamin"
date: "12/17/2020"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Chapter 2 The Very Basics

## 2.3 Functions
`round` function is used to round a number.
```{r}
round(3.1415)
```
`factorial` function is used to calculate the factorial of a number.
```{r}
factorial(3)
```
`mean` function is used to compute the mean of number(s).
```{r}
mean(1:6)
```
We can combine `round` and `mean` functions as follows:
```{r}
round(mean(1:6))
```
The `sample` function is used to sample.
```{r}
die <- 1:6
sample(x = die, size=1)
```
`args` can be used to view the arguments of a function.
```{r}
args(round)
args(sample)
```
`round` method can be accompanied with `digits`, for example:
```{r}
round(3.1415, digits = 2)
```
By **default**, the `sample` function is _without replacement_.
```{r}
sample(die, size=2)
```
When we want to `sample` _with replacement_, we can set `replace=TRUE` as follows:
```{r}
sample(die, size=2, replace = TRUE)
```
If you want to add up the dice, you can feed your result straight into the `sum` function:
```{r}
dice <- sample(die, size=2, replace = TRUE)
dice

sum(dice)
```

## 2.4 Writing Your Own Functions
We are going to write a function which returns the sum of rolling two dice.
```{r}
roll <- function(){
  die <- 1:6
  dice <- sample(die, size=2, replace = TRUE)
  sum(dice)
}
result <-  roll()
result
```
When you run a function in R, R will execute all of the code in the body and then return **the result of the last line of code**.
