---
title: "Chapter 2 Single-parameter Models"
author: "Hendra Bunyamin"
date: "12/20/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Let's import ggplot2 for plotting.
```{r}
library(ggplot2)
theme_set(theme_minimal())
```

## Probability of a girl birth given placenta previa 
437 girls and 543 boys have been observed. Calculate and plot the posterior distribution of the proportion of girls $\theta$, using uniform prior on $\theta$.

### Analysis using a uniform prior distribution
Under a uniform prior distribution for the probability of a girl birth, the posterior distribution is Beta(438, 544).    
`seq` creates evenly spaced values as follows:
```{r}
df1 <- data.frame(theta = seq(0.375, 0.525, 0.001))
a <- 438
b <- 544
head(df1)
```

As we read page $35$ from the book, the posterior density for $\theta$ is 
$$ \begin{aligned}
  \Pr(\theta \mid y) &\propto \theta^y (1 - \theta)^{n-y} \theta^{\alpha-1} (1-\theta)^{\beta - 1} \\
                     &= \theta^{y+\alpha-1} (1-\theta)^{n-y+\beta-1} \\
                     &= \text{Beta}(\theta \mid \alpha+y, \beta+n-y).
\end{aligned} $$

Specifically, `dbeta` computes the posterior density.   
```{r}
df1$p <- dbeta(df1$theta, a, b)
head(df1)
```

Next, we compute $95\%$ **central interval**.    
The `seq` creates evenly spaced values from $2.5\%$ quantile to $97.5\%$ quantile (i.e., $95\%$ central interval).     
The `qbeta` computes the value for a given quantile given parameters `a` and `b`.

```{r}
df2 <- data.frame(theta = seq(qbeta(0.025, a, b), qbeta(0.975, a, b), length.out = 100))
```

Subsequently, we compute the _posterior density_.     
```{r}
df2$p <- dbeta(df2$theta, a, b)
```

We also plot the posterior ($\text{Beta}(438, 544)$) and $48.8\%$  line for population average.
```{r}
ggplot(mapping = aes(theta, p)) +
  geom_line(data = df1) +
  # Add a layer of colorized 95% posterior interval
  geom_area(data = df2, aes(fill='1')) +
  # Add the proportion of girl babies in general population
  geom_vline(xintercept = 0.488, linetype='dotted') +
  # Decorate the plot a little
  labs(title='Uniform prior -> Posterior is Beta(438,544)', y = '') +
  scale_y_continuous(expand = c(0, 0.1), breaks = NULL) +
  scale_fill_manual(values = 'lightblue', labels = '95% posterior interval') +
  theme(legend.position = 'bottom', legend.title = element_blank())
```

