---
title: "Chapter 8 File Processing"
author: "Hendra Bunyamin"
date: "2023-07-28"
output: pdf_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Investigating Files

Pandang file, `intro.txt` yang berisi

`A computer program is a sequence of commands and instructions to effectively solve a given problem. Such a problem may involve calculations, data processing, or both. Each computer program is based on an underlying procedure called algorithm. An algorithm may be implemented in different ways, leading to different programs using the same procedure. We follow this convention throughout this book, where an algorithm refers to a list of procedures whereas a program refers to its implementation as a code.`    
   
```{r countwords}
countwords = function(){
  v = scan("../datasets/intro.txt", "", quiet = TRUE)
  print( paste( "The file has", length(v), "words." )  )
}
# Mari panggil fungsi countwords
countwords()
```
   



```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
